Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-04-20 doc/invt.tex
edit abstract

\title{Optimal updating factor in Wang-Landau and metadynamics simulations}

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing
  \note{``incrementally building up''?}%
  a bias potential to offset
  the potential of mean force (PMF).
  %
  The two differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bias potential of
  the bin containing the current $z$,
  while metadynamics updates also
  those of neighboring bins according to
  a Gaussian window.
  %
  The error of the final bias potential,
  hence that of the PMF,
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a method of computing the optimal schedule
  for a general multiple-bin updating scheme.
  %
  We show that the optimal schedule of metadynamics
  depends on the simulation length
  and the width of the Gaussian.
  %is solved from the equation of motion
  %of a free particle with a position-dependent mass.
  %
  Further,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of multiple-bin bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  The optimal schedule of these schemes
  is given by the inverse-time formula,
  with a shifted origin of time such that
  the initial updating magnitude is half of
  the previous equilibrium value.
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case,
and gives the optimal value of the origin of time.
%
It also predicts, for metadynamics,
a more complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the single-bin updating scheme
used by the WL algorithm
to a class of multiple-bin updating schemes
that are optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model; or the temperature index
in a simulated tempering simulation\cite{
marinari1992, lyubartsev1992}.
%
For a continuous quantity, $z$,
which is typical in a molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \delta z)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation.
%
Adaptive FDS methods,
including the WL algorithm and metadynamics,
effectively sample a flat distribution
by constructing a bias potential
to offset the PMF.
%
The optimal schedule ensures fastest convergence
of the bias potential.
%They differ by the updating schemes,
%which specify how the bias potential is modified
%in each updating step.
%%
%In the single-bin or WL case,
%the update is limited to the current bin;
%while in the metadynamics case,
%the bias potential of a neighborhood of the current one
%is updated with relative weight specified by a Gaussian window.

Each adaptive FDS method defines an updating scheme,
or mathematically an updating matrix,
whose elements $w_{ij}$ specifies the
relative magnitude of updating the bias potential
at bin $i$ upon a visit to bin $j$.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
and the integrals of the correlation functions
the corresponding eigenmodes.
%
For the single-bin (WL) scheme,
the optimal schedule from our method
is given by inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
The parameter, $t_0$, can be adjusted
according to Eq. \eqref{eq:t0_sinc},
to accommodate the initial error.
%such that the initial updating magnitude
%is half of the previous equilibrium value.
%
For a general multiple-bin scheme,
including the Gaussian one used in metadynamics,
the optimal schedule has no closed form,
but is given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the local MC simulation.
    %
    The MD simulation was ran for $10^{10}$ steps with
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$,
    and the temperature was $1.0$.
    %
    The velocity rescaling\cite{bussi2007} thermostat was
    applied every MD step with the viscocity set to 20.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      Thee data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

>>>>

\title{Optimal updating factor in adaptive flat-distribution-sampling simulations}

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for adaptive flat-distribution sampling methods,
  such as the Wang-Landau (WL) algorithm and metadynamics.
  %
  These sampling methods
  construct a bias potential to offset
  the potential of mean force
  by incremental updating loops.
  %
  The updating magnitude should decrease over time
  to reduce the fluctuation of the bias potential,
  and the optimal schedule optimizes
  the rate of convergence.
  %
  We derive the optimal schedule from
  an equation of motion
  of a free particle with a position-dependent mass,
  %
  and show that
  the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  we show that the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is given by a modified inverse-time formula.
  %
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for a general adaptive flat-distribution sampling methods,
including those used in WL and metadynamics simulations.
%
In this method, the optimal schedule is derived
from an equation of motion of a free particle,
with the initial updating magnitude set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
the inverse-time formula,
with the origin of time modified
to satisfy the above optimal initial condition.
%
For metadynamics, however,
the optimal schedule is more complex,
and it can be sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for infinitely long simulations,
and the modified inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model; or the temperature index
in a simulated tempering simulation\cite{
marinari1992, lyubartsev1992}.
%
For a continuous quantity, $z$,
which is typical in a molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \Delta z)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation,
e.g., the WL algorithm or metadynamics.
%
Adaptive FDS methods
effectively sample a flat distribution
by incrementally updating the bias potential
to offset the PMF.
%
The optimal schedule ensures fastest convergence
of the bias potential.
%They differ by the updating schemes,
%which specify how the bias potential is modified
%in each updating step.
%%
%In the single-bin or WL case,
%the update is limited to the current bin;
%while in the metadynamics case,
%the bias potential of a neighborhood of the current one
%is updated with relative weight specified by a Gaussian window.

Each adaptive FDS method is associated with
an updating scheme,
or mathematically an updating matrix,
whose elements $w_{ij}$ specifies the
relative magnitude of updating the bias potential
at bin $i$ upon a visit to bin $j$.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
and the integrals of the autocorrelation functions
of the eigenmodes.
%
For the single-bin (WL) scheme,
the optimal schedule from our method
is given by inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
The parameter, $t_0$, can be adjusted
according to Eq. \eqref{eq:t0_sinc},
to accommodate the initial error.
%such that the initial updating magnitude
%is half of the previous equilibrium value.
%
For a general multiple-bin scheme,
including the Gaussian one used in metadynamics,
the optimal schedule has no closed form,
but is given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the local MC simulation.
    %
    The MD simulation was ran for $10^{10}$ steps with
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$,
    and the temperature was $1.0$.
    %
    The velocity rescaling\cite{bussi2007} thermostat was
    applied every MD step with viscocity 20.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      Thee data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}
